Augmentation of a Historical Harpsichord Keyboard Replica for Haptic-Enabled Interaction in Museum Exhibitions

### 2025-01-22 src/article.tex
Updates from Overleaf

Technological advances have transformed how museums document, present and interpret their collections, fostering immersive experiences through tools such as laser scanning, 3D printing, and virtual reality \cite{allard2005use,Wachowiak01082009,RCM_2024_3D,Kuzminsky_LaserScan_2012,Schaich_3D_2007}. These technologies emphasise experiential authenticity, enabling encounters that evoke the past's sensory, emotional, and intellectual essence \cite{trant_Auth_1999}. However, as Pine and Gilmore note \cite{pinegilmore_2007}, achieving authenticity requires museums to navigate the delicate balance between preservation and meaningful engagement—a challenge that is particularly evident in the case of historical musical instrument collections \cite{McAlpine2014}.

Musical instruments represent a distinctive fusion of form, function, and history. Their cultural value extends beyond their visual appeal to include the tactile and auditory dimensions of use \cite{Fritz2017}. Yet, preservation concerns often limit direct interaction, reducing these artefacts to static displays. This ``red velvet cord'' approach, as theorised by McAlpine \cite{McAlpine2014}, protects fragile mechanisms but diminishes the instruments’ functional identity, disconnecting visitors from the full richness of their historical and cultural context.

The Tagliavini Collection in Bologna \cite{Tagliavini2007}, renowned for its historical keyboard instruments, exemplifies this dilemma. To address it, this project introduces an augmented replica of a historical harpsichord keyboard. 
% The project builds on an existing design borrowed from \cite{McPherson2013} and aligns with principles outlined in \cite{Masu_NIME_2023}, emphasising the importance of extending the lifecycle of existing NIMEs. 
The project fosters and complements the augmented piano keyboard design presented by McPherson \cite{McPherson2013} and ensures the relevance of previous NIME designs beyond their original context \cite{Masu_NIME_2023}.

@@

McAlpine discusses a case similar to the Tagliavini collection in his examination of the Benton Fletcher Collection at National Trust Fenton House \cite{McAlpine2014}. When these instruments were donated, Benton Fletcher stipulated that they remain playable and should continue to be maintained for tuition and public performance. A large sampling campaign was conducted, and a custom MIDI interface was designed to fulfil this requirement while preserving the original instruments' integrity. The MIDI keyboard, comprising two commercially available keyboards mimicking the two-manual harpsichord layout, was used by visitors to trigger the instrument samples recorded with tailored strategies for each. However, user tests identified a significant limitation: the commercially available weighted keys failed to provide an authentic sense of interacting with a historical keyboard \cite{McAlpine2014}.

@@

This work inherits the same philosophy as the Tromba Moderna project. By augmenting a replica of a historical harpsichord keyboard using minimally invasive electronics and controlling a MIDI-triggered sample library, the project intends to offer a tool enhancing the fruition of the Tagliavini collection whilst retaining a form of continuity with historical instrument-building traditions. Furthermore, the electronics design borrows ideas from a previous NIME by McPherson \cite{McPherson2013}. As such, the interface proposed here is not, strictly speaking, a \emph{new} musical interface, particularly in its haptic response designed to adhere to longstanding harpsichord building traditions. However, this work complements and follows up on previous reflections regarding the \emph{sustainability of results} within the NIME community and identified as \emph{the O in NIME} \cite{Masu_NIME_2023}. The intended use of the keyboard through meaningful interaction with a museum exhibit is where the novelty of this work lies, rather than solely in its technological development.

Beyond enhancing the visitor experience, future design iterations will serve as a research tool to explore the unique characteristics of the harpsichord and its impact on performance through the \anon{ERC-funded NEMUS} project, aiming to virtually reproduce the sound of historical keyboard instruments, and upcoming projects such as Rem@ke \cite{remake1}.

@@

\subsection{Open-source considerations}
A further objective, set forth by the authors, was to ensure accessibility and reproducibility by committing to an \emph{open-source} approach for all hardware, software, and data. The commitment to open-sourcing encompassed all aspects of the system, including hardware schematics, firmware, and calibration data. Cost-effectiveness was also a central consideration. The system was initially developed for a 3-key prototype and successfully scaled to 49 keys without significant increases in cost or complexity, and was designed to be easily built using resources typically available in a university-managed makers space.

Key components, such as QRE1113 sensors and CD4051BE multiplexers, were selected for their affordability and availability, while the modular PCB design ensured easy replication and maintenance.

The Arduino Nano BLE was chosen as the core microcontroller for its compatibility with open-source tools and ability to support both USB and BLE MIDI. Ferroelectric RAM (FRAM) was employed for data storage, offering an economical yet robust solution for preserving calibration settings across power cycles. Calibration workflows were optimised using the Arduino IDE’s serial plotter and open-source MIDI Monitor software, reducing reliance on proprietary tools and simplifying the process for users.

@@

The hardware design aimed to fulfil the earlier constraints, including non-invasiveness, reliability, and scalability. To achieve these goals, the system evolved through iterative prototyping, beginning with simple threshold-based testing and culminating in a fully functional multi-sensor interface capable of triggering MIDI events.

@@

The initial stage of development focused on testing whether sensor data could reliably trigger MIDI playback. Modifying an existing harpsichord for testing was considered but ultimately discarded due to significant internal measurement and layout discrepancies. Instead, a custom 3-key harpsichord mechanism (Figure \ref{fig:3key}) was acquired from the museum and used as a foundation for prototyping. This approach followed a methodology similar to that used in Timmermans' Haptic Key project \cite{Timmermans2020}, enabling incremental validation of hardware and sensor integration.

The 3-key model facilitated iterative testing of individual components, including sensor placement, signal processing, and mechanical tolerances. The model’s compact form allowed for rapid experimentation while preserving the physical properties of a full-scale harpsichord.

@@

% \begin{figure}
%     \centering
%        \begin{subfigure}
%          \centering
%          \includegraphics[width=\linewidth]{src/images/CrossSectionSensorPlacement.jpg}
%          \caption{Cross Section of 49-Key Model}
%          % \label{fig:y equals x}
%      \end{subfigure}\\
%     \begin{subfigure}
%          \centering
%          \includegraphics[width=\linewidth]{src/images/3-key-side-labelled.png}
%          \caption{3-Key Model: A: Wrestplank, B: Key Lever, C: Key Cap, D: Soundboard, E: Jack}
%          % \label{fig:y equals x}
%      \end{subfigure}        
%     \caption{Cross Sections of Models}
%     \label{fig:3key}
% \end{figure}

@@

To ensure the system's suitability for a museum context, the following criteria were established to guide sensor selection and integration:

\begin{itemize}
    \item \emph{Non-invasive:} Sensors must not require significant modifications to the harpsichord's mechanical structure or aesthetics.
    \item \emph{Low Latency:} Sensor response times were required to remain under 10 ms, with an upper limit of 25 ms deemed acceptable \cite{Jack2016}.
    \item \emph{Reliability:} Sensor data had to be accurate, minimising false positives and negatives without excessive filtering.
    \item \emph{Scalability:} The design needed to be cost-effective and adaptable for up to 50 or more keys while maintaining consistent performance.
    \item \emph{Expandability:} The system should accommodate future functionality, such as additional MIDI parameters or data visualisation.
\end{itemize}

These criteria ensured that the design aligned with the open-source ethos of the project, allowing assembly in standard maker spaces without specialised equipment.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{src/images/system-block-diagram-long.png}
    \caption{Block diagram of PCB connection. Sensor and LED are routed through each sensor PCB}
    \label{fig:system-block-diagram}
\end{figure*}

@@

The QRE1113 sensors function by detecting infrared light reflected from nearby surfaces. The gradient stickers affixed to each jack provided a consistent reflective surface for the sensors, enabling precise tracking of jack displacement. This approach was chosen for its simplicity and scalability, allowing the design to progress from the 3-key prototype to a 49-key model without significant modifications.

To eliminate cross-talk between adjacent sensors, 3D-printed baffles were installed on the PCBs (Figure \ref{fig:baffles}). These baffles, fabricated from dark-pigmented PLA, ensured that infrared reflections from neighbouring jacks did not interfere with sensor readings. Testing confirmed that this measure significantly improved data reliability.

\begin{figure}  
  \centering
  \includegraphics[width=\linewidth]{src/images/simple-schematic-bw-.jpg} 
  \caption{Optical sensor in a simple voltage divider circuit. \texttt{V\_OUT} is routed to one of 8 channels on the CD4051BE multiplexer}
  \Description{} 
  \label{fig:simple-schematic}
\end{figure}

\subsection{Controller Board}\label{controller-board}

The controller board, designed around the Arduino Nano BLE platform, was the central hub for processing sensor data and triggering MIDI messages. The Nano BLE was selected for its compact form factor, integrated BLE MIDI functionality, and native USB support. These features simplified development while ensuring compatibility with existing MIDI-based synthesis systems.

@@

Each sensor's threshold was manually adjusted using RGB LEDs integrated into the system. These LEDs provided visual feedback during calibration, enabling the identification of malfunctioning sensors and simplifying the alignment process. The calibration workflow utilised the Arduino IDE serial plotter and open-source MIDI monitoring software (Figure \ref{fig:serial_monitor}).

@@

@@

The assembly process at the NEMUS lab in Bologna involved meticulous adjustments to key clearances and wiring layouts. Modifications included milling sections of the keys to ensure adequate clearance for the PCBs and varnishing the jacks to improve sticker adhesion. Subsequent iterations addressed these challenges, refining the system’s scalability and robustness.

% \begin{figure}
%     \centering
%     \includegraphics[width=\linewidth]{src/images/CrossSectionSensorPlacement.jpg}
%     \caption{Cross-section illustrating sensor placement relative to the jack mechanism.}
%     \label{fig:cross-section}
% \end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/images/sensor-board-w-baffles.jpeg}
    \\
    \includegraphics[width=\linewidth]{src/images/baffles.png}
    \caption{Baffles designed to prevent cross-talk between adjacent sensors.}
    \label{fig:baffles}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/images/serial_monitor.png}
    \caption{Arduino IDE serial plotter used for calibration of sensor thresholds. \texttt{M} and \texttt{m} represent the maximum an minimum value of digitised data. \texttt{K} and \texttt{T} are the values of the sensor and threshold being calibrated, followed by a two digit number denoting the sensor index (in this case sensor 32).}
    \label{fig:serial_monitor}
\end{figure}

% \section{Hardware Design}

% For the first iteration the data from the sensor would simply be used to test when a threshold had been passed. On passing the threshold, a message will be sent to trigger the playback of a corresponding note.

% For the initial prototyping stage, modifying an existing harpsichord was considered though ultimately discarded. An existing instrument would
% have provided a test for scaling the electronics, but internal measurements and layout would have been too different.

% Using a similar approach to Timmermans' Haptic Key project, \cite{Timmermans2020}, our first step was to fabricate or procure a model of a single working key. San Colombano provided a model 3-key of a harpsichord mechanism by Graziano Bandini (Figure \ref{fig:3key}).

% \subsection{Sensor Criteria}\label{sensor-criteria}

% A set of criteria was drawn to which the final sensor system would have to satisfy. These criteria were used to evaluate the feasibility throughout the prototyping phase.

% \begin{itemize}
% \item
%   Non-invasive: The sensor system should not require major modifications
%   to the interface on which it is installed.
% \item
%   Low Latency: Reading sensors and then triggering an output should have
%   a low latency. Following \cite{Jack2016} an latency was set as 10 ms
%   with an upper limit set to 25 ms beyond which the sensors would be
%   deemed unusable
% \item
%   Reliable: Data obtained should be reliable and accurate enough not to
%   result in false positives or false negatives without extensive
%   filtering
% \item
%   Scalable: The system should be scalable both economically and
%   temporally. Given the open source nature of the project, it could not
%   be prohibitively expensive to carry out. Also any system that would
%   work on a single key would have to scale financially to potentially 50
%   or more. The time required to install and calibrate the sensors would
%   also have to scale.
% \item
%   Expandable: Any sensor system should be functionally expandable to
%   allow for the usage of the data for control over other parameters.
% \end{itemize}

% The sensor system was designed so that they would not require any equipment outside what can commonly be found in a maker space at a university for assembly. This decision was motivated by the project's goal to be open source and reasonably recreated without procuring additional specialist equipment.

% The electronics system can be divided into two designs, the "sensor board" (Figure \ref{fig:7-sensor-board}) for components for detecting jack displacement and controller board for collecting and processing sensor signals.

% \begin{figure}  
%   \centering
%   \includegraphics[angle=-90,width=\linewidth]{src/images/7-sensor-board-v0.1.0.png} 
%   \caption{} 
%   \Description{} 
%   \label{fig:7-sensor-board}
% \end{figure}

% \subsection{Sensor board}\label{sensor-board}

% The 49 optical sensors are divided across 7 PCBs each containing

% \begin{itemize}
% \item
%   x7 QRE1113 optical sensors
% \item
%   x7 100 $\Omega$ resistors
% \item
%   x7 10 k$\Omega$ resistors \footnote{a update on the design showed these could be reduced to a single resistor.}
% \item
%   x1 CD4051BE Multipexer
% \end{itemize}

% A number of sensors were tried before the final system was decided upon. 
% Sensors tested for the design included, a 6-degrees-of-freedom inertial measurement unit, reed and hall sensors paired with magnets, and light-dependant resistors mounted to the jack rail.
% None of the above satisfied the sensor criteria described and were thus discounted.

% Following from the work in \cite{McPherson2013, McPherson2019}, we tested
% a system using the Fairchild QRE1113 \footnote{Fairchild, now ON Semiconductor, QRE1113 datasheet \url{https://www.onsemi.com/download/data-sheet/pdf/qre1113-d.pdf}}. The
% QRE1113 is a combination infrared LED and phototransistor sensitive to
% IR light. The phototransistor of QRE1113 provides data on how much light
% is present and in particular how much light is being reflected from a
% surface. Since reflected light will be proportional to the distance of
% a surface it is a good, close-proximity, distance sensor. Distance is
% the way in which the instruments in \cite{McPherson2013, McPherson2019}
% function as well the Moog Piano Bar from which it took inspiration.

% The limitation for taking a distance approach is that there is not one
% but 2 jacks per key. For each jack to be measured independently the data
% would need to come from the jacks themselves.

% % They \emph{could} be used and individual trigger points set in
% % calibration, but this creates new problems.

% \begin{figure}  
%   \centering
%   \includegraphics[,width=\linewidth]{src/images/CrossSectionSensorPlacement.jpg} 
%   \caption{} 
%   \Description{} 
%   \label{fig:cross-section}
% \end{figure}

% The optical sensors were directed at a greyscale gradient printed on a sticker and applied to the side of the jack.
% This was first trialled using a generic inkjet printer and double-sided tape. 
% For the final iteration, the stickers were printed on Coala 1D 100 Gloss PG 100 $\mu$m white gloss monomeric PVC with permanent grey acrylic dispersion adhesive \footnote{Datasheet: \url{https://www.antalis.co.uk/mediashare/g4media/pdf/TS_EN_COALA_1D_100_GLOSS_P_00_ISS_15022023.pdf}} using a HP Latex 115 and then cut with a Summa 150 roll cutter. 
% The printer and roll cutter greatly reduced the manual labour for cutting stickers meaning the process would scale from the 6 jacks of the 3-key to the 98 jacks of the 49-key model.

% \begin{figure}  
%   \centering
%   \includegraphics[angle=90,width=\linewidth]{src/images/jack-w-tag.jpeg} 
%   \caption{} 
%   \Description{} 
%   \label{fig:jack-w-tag}
% \end{figure}

% During testing it was found that their was a considerable amount cross-talk between adjacent sensors. 
% This meant the reading from key would be affected by the movement of the key next to it. 
% The cross-talk was assumed to be as a result of light reflecting from other jacks.
% Given that IR is used this was difficult to confirm by sight. 
% A set of baffles was designed for the PCB that would wrap round the jack and stop light from other jacks or sensors \ref{fig:baffles}. 
% The baffles were printed on a BAMBU X1 3D printers using PLA plastic. A dark pigmented PLA was used to avoid IR light simply being diffused. 
% After the baffles were fitted, the cross-talk was eliminated.

% \begin{figure}  
%   \centering
%   \includegraphics[,width=\linewidth]{src/images/baffles.png} 
%   \caption{} 
%   \Description{} 
%   \label{fig:baffles}
% \end{figure}

% Calibration to pluck thresholds were to be carried out manually.  To allow for easy identification of the sensor being addressed during the calibration process, RGB LEDs were integrated into the design. The LEDs were also used in the identification of any malfunctioning or un-calibrated sensor.

% TLC5940 or other PWM driver such as in \cite{McPherson2013} add robustness should an LED fail, but complexity to assembly process. 
% Instead, the design uses addressable LEDs with integrated driver that share the same data line. 
% This has the limitation where a faulty LED will stop all subsequent LEDs in the line from working.

% The BLE Nano has 8 ADC pins connected to one ADC channel via a multiplexer. 
% We avoided the complexity of communicating multiple microcontrollers and instead used a Texas Instruments CD4051B multiplexer on each sensor board.
% An 8-channel multiplexer was chosen as the model has 49-keys and a division of 7 PCBs each containing 7 sensors.
% The multiplexer reduced the 7 sensor signals to a single channel. The CD4051B is addressed with a 3-bit which necessitates 3 digital channels.
% Transistor / Op-amp circuit could have been used similar to the one in \cite{McPherson2013}.

% \subsection{Controller Board}\label{controller-board}

% Controller board designed around Arduino nano form factor. The Nano is
% connected through headers to allow easy switching between chips. Should
% another chip in a Nano for factor be found it could supplant the current
% Nano without any need for a redesign.

% The first version of the board exclusively used 2.54 mm pitch headers.
% The reasons for this were flexibility when creating cable looms and to
% leave open the possibility of rewiring the PCB.

% This was the case for the LEDs as they initially were to be powered on
% 5V. It was discovered that the 3V regulator of the nano was sufficient
% to power all LEDs at a lower brightness while reducing current draw.
% Since brightness was not important.

% The 2.54 mm pitch headers had two major drawbacks. First, the orientation
% of the cables could be reversed, which in the best case scenario would
% simply mean the system did not work correctly and in the worst case it
% could potentially damage the components.

% Secondly, the headers do not connect securely and wait of the cabling
% was enough for components to become disconnected as the harpsichord was
% handled.

% To address this, once the wiring for the controller board was finalised
% a new design was drafted this time using JST-PH connectors. No
% commercial cables were immediately available in the configuration and
% length required, as such a cable loom needed to be made which meant
% crimping cables by hand. A crimping tool from JST is prohibitively
% expensive compared to the cost of other components in the project
% \footnote{from RS \url{https://uk.rs-online.com/web/p/crimp-tools/6880877}
% £508 (inc. VAT)}. The crimping tool also appeared to be uncommon and was
% not readily available in the workshops at the project's disposal.
% Eventually a tool was found, but an alternative approach was also
% researched in order to maintain the criteria of accessibility.

% A locator was adapted from a creative commons STL model \footnote{\url{https://www.thingiverse.com/thing:1646016}} that allowed a generic crimping tool to be adapted for use with JST-PH crimps.

% \subsubsection{Arduino Nano BLE}\label{arduino-nano-ble}

% Arduino Nano BLE was chosen for a few reasons. The Nano form factor was
% appealing as it would allow for easy changes between chipsets without
% requiring any rewiring and thus no new PCBs would need to be created if
% a better alternative was found later in the project. Another benefit to
% the Arduino Nano boards was that typically had native-USB meaning the
% could be programmed as a hardware USB MIDI device directly.

% The Nano BLE was used in the initial stages for its integrated inertial measurment unit.

% When it was finalised that the project would proceed using the QRE1113 a
% number of microcontroller units (MCU) were considered.

% The MCUs considered were the Arduino Nano 33 IoT, Arduino Nano 33 BLE,
% Arduino Nano ESP32 and STM32 NUCLEO-L031K6.

% A rough performance test was later carried out using an STM32
% NUCLEO-L031K6, a Nano ESP32 and the BLE. The Nano BLE was found to be
% more performative. To execute a single sensor read cycle of the firmware
% the STM32 took 40 ms, and both the Nano ESP32 and Nano BLE required 11 ms
% on average. The Nano ESP32 had high variability so it was decided to
% continue with the Nano BLE.

% BLE functionality also meant if the cable connection between the nano
% and computer running audio synthesis was impractical, BLE MIDI was
% available as a fallback. The Nano BLE also has a 12-bit DAC, which
% provided a contingency should the default 10-bit not be sensitive enough
% for the signal from the sensors.

% Initially a Nano IoT was used for both WiFi and Bluetooth Low Energy
% functionality, but it was found that 2 ADC channels were in fact not
% useable and would have required additional multiplexers.

% \subsubsection{EEPROM / NVRAM / FRAM}\label{eeprom-nvram-fram}

% There would need to be the ability to store values on non-volatile
% memory so that the electronics could be powered-down without losing
% data.

% Adjusting the values through firmware would not have been flexible or
% practical. Instead an Ferroelectric RAM unit was chosen and connected
% using an SPI interface. NVRAM or EEPROM would have sufficed and the FRAM
% was primarily chosen for it's ready availability.

% Most microcontrollers have some variety of non-volatile program memory
% that can be read and writ accessed but in most cases this memory is
% cleared during compilation and re-flashing. The small cost was far
% outweighed by the potential for data loss if using onboard RAM.

% SD card would provide an easier way to interact with calibration data.
% The position of the controller board did not permit easy access.
% Reliability and speed was preferred over storage capacity.

% \subsection{Power}\label{power}

% All electronics combined draw around 1.1 A of current at 5V with some fluctuation during start-up.

% Arduino BLE has some functionality disabled to limit current draw.

% Ideally the unit could be bus powered from USB.

% The sensors are in a state of constantly drawing power. 
% It is possible that powering only when data is required could be reduce power enough.

% Current designs have a separate MIDI device for each jack row. 
% Combining the devices and also reducing current draw to below the common 0.5 amps for a USB port would certainly be a challenge.

% \section{Firmware Design}\label{firmware-design}

% Arduino platform chosen for wide adoption, relative ease of use and
% library support for components.

% Firmware checks components are connected then reads the FRAM to see if
% threshold data is already present. The rotary encoder and it's momentary
% switch are polled for any change in state. A key is selected with the
% rotary encoder while the momentary switch is used for changing between
% states for key selection and threshold adjustment A double click of the
% rotary encoder writes the current threshold values to FRAM. The sensors
% for the current multiplexer channels are polled. If any sensor reads
% above the threshold when it was previously below a MIDI On message is
% sent. For the reverse case where the value was below the threshold when
% it was previously above, a MIDI Note Off message is sent.

% \section{Implementation}\label{implementation}

% \begin{figure}  
%   \centering
%   \includegraphics[,width=\linewidth]{src/images/exhibition-user-1.jpeg} 
%   \caption{} 
%   \Description{} 
%   \label{fig:exhibition}
% \end{figure}

% \subsection{Installation}\label{installation}

% The electronic components were partly pre-assembled at the university of
% Edinburgh. Though all measurements were known ahead, enough flexibility
% was needed to allow for problems to solved in situ.

% The final assembly took place at the NEMUS lab at the University of
% Bologna over the final week in October 2024 for delivery at San
% Colombano for exhibition in the first week of November in 2024.

% A misreading in the original clearance between the keys and the PCB
% meant that the keys were blocked entirely from moving. A 10 mm by 15 mm
% section was milled out of each key to create extra
% clearance. The milling required the removal, shortening and refitting of
% the leather pads for the jacks. The extra work
% added an extra day to the full install.

% \begin{figure}  
%   \centering
%   \includegraphics[width=\linewidth]{src/images/milled-keys.jpg} 
%   \caption{} 
%   \Description{} 
%   \label{fig:milling}
% \end{figure}

% A redesign of the PCBs meant that removing the material would no longer
% be necessary.

% The PCBs shared power and data lines for components with the exception
% of the output from the multiplexer. Ribbon cable was cut to size to
% allow for some room to readjust the distance of the sensors from the
% jacks. The data signals from each PCB were connected using a separate
% cable loom. Soldering cables and inspecting each join required another
% day work. This did allow for readjustments to the position of the
% controller board Measurements of the rear chamber did not match with the
% Mac Mini that was to be used a different layout was decided for fitting
% the controller board/ Cutting cables to size allowed for a great deal of
% flexibility but at the cost of time. After the first install it was
% clearer what constraints were in place for fitting and an alternative
% wiring system using flat flex cable was implement (discussed in
% \textbackslash section)

% Fitting PCBs required the removal of all keys. The support brackets were
% assembled and screwed into the the underside. With the harpsichord on
% it's back the sensors were tested to ensure an expected range.
% Self-tapping round head screws were used to to attach the brackets. The
% screws were not over tightened to allow for brackets to move. The
% brackets were not easily adjusted once the key were replaced. Adjusted
% to a best average with subsequent adjustment carried out on the
% microcontroller.

% As the PCBs were fitted the jacks had the gradient stickers applied to
% them. The jacks had been coated with a varnish on the short side to
% provide a better surface for the stickers to adhere to. After the
% stickers were applied excess material was trimmed from each jack. The
% jack were then check that they moved freely in their slot.

% After fitting the PCBs preliminary calibration was carried out. Before
% the keys were refitted the jacks were put back in place. The data from
% each key was confirmed to be in a similar range. Any key showing
% anomalous data had it's jack removed and re-inspected. The PCB fitting
% process and refitting required another day as faults were found in
% wiring or misalignment of the stickers. After the sensors were
% confirmed to be functioning the keys were put back.

% The calibration workflow was with \anon{Dr. Craig Webb} who was unfamiliar with
% the system. Adjustments were made to the firmware to to make it more
% intuitive for the user. Addressable LEDs became vital as it allowed for
% easy identification of the key being calibrated. They were also used to
% identify any keys whose current reading was beyond what was expected.

% The rotary encoder was used as the interface to select a key, adjust
% it's threshold and update the Ferroelectric RAM (FRAM) with new values.

% A process was implemented where the key was selected by playing, but
% this caused problems in situations where the data was unusual due to
% misalignment of the sensors.

% This would likely be a better system for when the keyboard was fully
% operational.

% \begin{figure}  
%   \centering
%   \includegraphics[width=\linewidth]{images/serial_monitor.png} 
%   \caption{} 
%   \Description{} 
%   \label{fig:serial_monitor}
% \end{figure}

% The thresholds were calibrated visually utilising the serial plotter of
% the Arduino IDE (Figure \ref{fig:serial_monitor}) and the open source MIDI
% Monitor software \footnote{\url{https://github.com/krevis/MIDIApps}}.

% \begin{itemize}
% \item
%   partial pre-assembly
% \item
%   handsolder components

>>>>

Technological advances have transformed how museums document, present and interpret their collections. Immersive experiences are afforded through tools such as laser scanning, 3D printing, and virtual reality \cite{allard2005use,Wachowiak01082009,RCM_2024_3D,Kuzminsky_LaserScan_2012,Schaich_3D_2007}. These technologies form a kind of experiential authenticity, enabling encounters that evoke the past's sensory, emotional, and intellectual essence \cite{trant_Auth_1999}. However, as Pine and Gilmore note \cite{pinegilmore_2007}, achieving authenticity requires museums to navigate the delicate balance between preservation and meaningful engagement—a challenge that is particularly evident in the case of historical musical instrument collections \cite{McAlpine2014}.

Musical instruments represent a peculiar fusion of form, function, and history. Their cultural value extends beyond their visual appeal to include the tactile and auditory dimensions of use \cite{Fritz2017}. Yet, preservation concerns often limit direct interaction, reducing these artefacts to static displays. This ``red velvet cord'' approach, as theorised by McAlpine \cite{McAlpine2014}, protects fragile mechanisms but diminishes the instruments’ functional identity, disconnecting visitors from the full richness of their historical and cultural context.

The Tagliavini Collection in Bologna \cite{Tagliavini2007}, renowned for its historical keyboard instruments, exemplifies this dilemma. To address it, this project introduces an augmented replica of a historical harpsichord keyboard. 
% The project builds on an existing design borrowed from \cite{McPherson2013} and aligns with principles outlined in \cite{Masu_NIME_2023}, emphasising the importance of extending the lifecycle of existing NIMEs. 
% The project fosters and complements the augmented piano keyboard design presented by McPherson \cite{McPherson2013} and ensures the relevance of previous NIME designs beyond their original context \cite{Masu_NIME_2023}.

@@

McAlpine discusses a case similar to the Tagliavini collection in his examination of the Benton Fletcher Collection at National Trust Fenton House \cite{McAlpine2014}. When these instruments were donated, Benton Fletcher stipulated that they remain playable and should continue to be maintained for tuition and public performance. A large sampling campaign was conducted, and a custom MIDI interface was designed to fulfil this requirement while preserving the original instruments' integrity. The MIDI controller, comprising two commercially available keyboards mimicking the two-manual harpsichord layout, was used by visitors to trigger the instrument samples recorded with tailored strategies for each. However, user tests identified a limitation: the commercially available weighted keys failed to provide an authentic sense of interacting with a historical keyboard \cite{McAlpine2014}.

@@

This work inherits the same philosophy as the Tromba Moderna project. By augmenting a replica of a historical harpsichord keyboard using minimally invasive electronics and controlling a MIDI-triggered sample library, the project intends to offer a tool enhancing the fruition of the Tagliavini collection whilst retaining a form of continuity with historical instrument-building traditions. Furthermore, the electronics design borrows ideas from a previous NIME by McPherson \cite{McPherson2013}. As such, the interface proposed here is not, strictly speaking, a \emph{new} musical interface, particularly in its haptic response designed to adhere to longstanding harpsichord building traditions. However, this work complements and follows up on previous reflections within the NIME community, emphasising the \emph{O} in NIME \cite{Masu_NIME_2023}. 
The intended use of the keyboard through meaningful interaction with a museum exhibit is where the novelty of this work lies, rather than solely in its technological development. Furthermore, relying on previous NIMEs, this project extends the longevity of the results beyond the scope envisaged in the original papers and enables a form of sustainability.

Besides enhancing the visitor experience, future design iterations will serve as a research tool to explore the unique characteristics of the harpsichord and its impact on performance through the \anon{ERC-funded NEMUS} project, aiming to virtually reproduce the sound of historical keyboard instruments, and upcoming projects such as Rem@ke \cite{remake1}.

@@

\subsection{Project Deployment}
A further objective, set forth by the authors, was to ensure accessibility and reproducibility by committing to an \emph{open-source} approach for all hardware, software, and data. The commitment to open-sourcing encompassed all aspects of the system, including hardware schematics, firmware, and calibration data. Cost-effectiveness was also a central consideration. The system was initially developed for a 3-key prototype, shown in Figure \ref{fig:3key}, and successfully scaled to 49 keys without significant increases in cost or complexity. Specifically, the system was designed to be easily assembled using resources typically available in a university-managed makers space.

Components, such as QRE1113 sensors and CD4051BE multiplexers, are widely available from commercial resellers, while the modular PCB design ensures easy replication and maintenance.

The Arduino Nano BLE was chosen as the core microcontroller for its compatibility with open-source tools and ability to support both USB and BLE MIDI. Ferroelectric RAM (FRAM) was employed for data storage since it offers an economical yet robust solution for preserving calibration settings across power cycles. Calibration workflows were optimised using the Arduino IDE’s serial plotter and open-source MIDI Monitor software, reducing reliance on proprietary tools and simplifying the process for users.

Reference repositories for this project can be found here. PLEASE FILL OUT AS APPROPRIATE (and anonymise).

@@

Figure \ref{fig:system-block-diagram} shows a block diagram of the finalised hardware setup. The system evolved through iterative prototyping, beginning with simple threshold-based testing and ending in a fully functional multi-sensor interface capable of triggering MIDI events.

@@

The initial stage of development focused on testing whether sensor data could reliably trigger MIDI playback. Modifying an existing harpsichord for testing was considered but ultimately discarded due to significant internal measurement and layout discrepancies. Instead, a custom 3-key harpsichord mechanism (Figure \ref{fig:3key}) was acquired from the museum and used as a foundation for prototyping. This approach followed a methodology similar to that used in Timmermans \emph{et al.}'s Haptic Key project \cite{Timmermans2020}, since the 3-key model enabled iterative testing of individual components, including sensor placement, signal processing, and mechanical tolerances, before upscaling.

@@

@@

The following criteria were established to guide sensor selection and integration to make the system suitable for a museum context:

\begin{itemize}
    \item \emph{Non-invasiveness.} Sensors must not require significant modifications to the harpsichord's mechanical structure or aesthetics.
    \item \emph{Low Latency.} Sensor response times were required to remain under 10 ms, with an upper limit of 25 ms deemed acceptable, following empirical criteria found in previous studies \cite{Jack2016}.
    \item \emph{Reliability:} Sensor data had to be accurate, minimising false positives and negatives without excessive filtering.
    \item \emph{Scalability.} The design needed to be cost-effective and adaptable for up to 49 keys while maintaining consistent performance.
    \item \emph{Expandability.} The system should accommodate future functionality, such as additional MIDI parameters or data visualisation.
\end{itemize}

The open-source ethos of the project ensured assembly was made possible in standard maker spaces without specialised equipment and provided further guidance in design choices.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{src/images/block-diagram.pdf}
    \caption{Block diagram of PCB connections. Sensor and LED are routed through each sensor PCB}
    \label{fig:system-block-diagram}
\end{figure*}

@@

The QRE1113 sensors detect infrared light reflected from nearby surfaces, as per the scheme in Figure \ref{fig:simple-schematic}. The gradient stickers affixed to each jack provided a consistent reflective surface for the sensors, which were used to track jack displacement precisely. This easily scalable system allowed straightforward progress from the 3-key prototype to a 49-key model without significant modifications.

\begin{figure}[b]  
  \centering
  \includegraphics[width=\linewidth]{src/images/simple-schematic-bw-.jpg} 
  \caption{Optical sensor in a simple voltage divider circuit. \texttt{V\_OUT} is routed to one of 8 channels on the CD4051BE multiplexer}
  \Description{} 
  \label{fig:simple-schematic}
\end{figure}

3D-printed baffles were installed on the PCBs to eliminate cross-talk between adjacent sensors, as per Figure \ref{fig:baffles}. These baffles, fabricated from dark-pigmented PLA, ensured that infrared reflections from neighbouring jacks did not interfere with sensor readings. Testing confirmed that this measure significantly improved data reliability.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/images/sensor-board-w-baffles.jpeg}
    \\
    \includegraphics[width=\linewidth]{src/images/baffles.png}
    \caption{Baffles designed to prevent cross-talk between adjacent sensors.}
    \label{fig:baffles}
\end{figure}
\subsection{Controller Board}\label{controller-board}

The controller board, designed around the Arduino Nano BLE platform, was the central hub for processing sensor data and triggering MIDI messages. The Nano BLE was selected for its compact form factor, integrated BLE MIDI functionality, and native USB support.

@@

Each sensor's threshold was manually adjusted using RGB LEDs integrated into the system. These LEDs provided visual feedback during calibration, enabling the identification of malfunctioning sensors and simplifying the alignment process. The calibration workflow utilised the Arduino IDE serial plotter and open-source MIDI monitoring software, as per Figure \ref{fig:serial_monitor}.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/images/serial_monitor.png}
    \caption{Arduino IDE serial plotter used for calibration of sensor thresholds. \texttt{M} and \texttt{m} represent the maximum an minimum value of digitised data. \texttt{K} and \texttt{T} are the values of the sensor and threshold being calibrated, followed by a two-digit number denoting the sensor index (in this case, sensor 32).}
    \label{fig:serial_monitor}
\end{figure}

@@

The assembly process at the NEMUS lab in Bologna involved meticulous adjustments to key clearances and wiring layouts. Modifications included milling sections of the keys to ensure adequate clearance for the PCBs and varnishing the jacks to improve sticker adhesion.

\section{Context}\label{context}

Designed as part of an exhibition for the museum San Colombano in Bologna. Instruments are to be played by the general public to interact with items in the collection that are no longer in playing condition. The audience of one who listens via headphones. The interface strings are damped with felt strips. The damping did change the tension and therefore the feel of the strings so
adjustments needed to be made to the tuning to accomodate for this.
